A Meta-Analysis of Music Emotion Recognition Studies

### 2025-03-21 manuscript/manuscript.qmd
changes during our last meeting

## The semantic gap in MER

@@

Our aim is to evaluate the predictive accuracy of two models of emotional expression in music: (a) models that predict track-specific coordinates in affective circumplex space (valence and arousal), and (b) models that classify discrete emotion categories. We focus on recent studies to identify the overall success rate in MER tasks and the key factors such as modeling techniques and features that contribute to the prediction accuracy of the models. To achieve this, we conduct a meta-analysis of journal articles published in the past 10 years. Based on existing literature, we hypothesize that arousal will be predicted with higher accuracy than valence, as valence tends to be more context-dependent and challenging to model. For emotion classification, we expect simple utilitarian emotions (e.g., fear, anger) will be easier to predict than more complex social emotions (e.g., sadness, nostalgia).

<!-- DISCUSS: we have not address this last thought at all, TE 19/3/2025 -->

@@

Figure 4 shows forest and funnel plot visualizations from the random-effects model of the best-performing models in classification studies. $MCC$s vary across a wide range, reporting $MCC$s between 0.55 to 0.98. Table 4 indicates that using the best model from each study increases performance relative to all models ($MCC$ = 0.8684 95%CI: 0.7475-0.9336), yet slightly increases heterogeneity ($\tau^2$ = 0.3180, $I^2$ = 99.8%).

@@

<! -- DISCUSS: 4 digits is too much, let's make this consistent, 3 in all places?, TE 19/3/2025 -->

>>>>

<!--paragraph on models used across regession & classification tasks-->
```{=html}
<!-- 
- Types of classification algorithms employed
- Popularity of SVMs (for both classification and regression)
- Algorithms used for personalized and nonpersonalized MER models?
- Success with regression varieties
- Introduction of neural networks & deep learning paradigms
-->
```

<!--summary of feature sets and quantity-->
```{=html}
<!-- 
- Wide set of features explored
- Summary of categories commonly used/discussed (Panda)
- Audio, midi, lyric features, tags 
- Low-, mid-, and high-level features
- Emphasize focus on audio-based featuers
- Small, medium, and large data sets (Features: 30, 300, 3000) 
- Could emphasize that types of features differ based on type of study
(psychology, engineering) due to different goals
-->
```

<!--MER approaches: psychology, engineering approaches-->
```{=html}
<!-- 
- Big picture: explanation vs predicting
- Summary of findings, success with MER approaches.
  - Scope of topic areas explored
  - Annotations: sometimes presented as a game, either experts or participants,
  sometimes secondary data from web databases (LastFM). Treated as "ground truth" (reference standard may be more accurate). Training/testing/validation. Comparisons with previously collected data/benchmarks
  - Participants: e.g., 5 experts, MTurk, etc.
  - Demographics
  - Common stimuli/genres: predominately pop
- Summary of findings, success with psychological approaches
  - Scope of topic areas explored
  - Annotations: Used to explain variance in participants' responses. Cross-validation rare.
  - Participants: psychology students, Prolific/MTurk?
  - Demographics
  - Common stimuli/genres: predominately classical
-->
```

## The semantic gap in MER / What are the reported standards? / ceiling
```{=html}
<!--
I'm considering now reworking this paragraph and de-emphasizing the semantic gap given the limited extent to which can address this. It might be worth including to show the semantic gap has narrowed, and we can point to the Barthet estimate, or to conclude that "linear methods with small to mid-sized feature sets appear perform best for this dimension".

- semantic challenge
- inherent error in all stages (annotation, features, models)

-->
```

@@

<!--This paragraph now is somewhat inconsistent with the analyses we actually perform motivating the journal/feature N/model discussion-->

Our aim is to evaluate the predictive accuracy of two models of emotional expression in music: (a) models that predict track-specific coordinates in affective circumplex space (valence and arousal), and (b) models that classify discrete emotion categories. We focus on recent studies to identify the overall success rate in MER tasks and the key factors such as modeling techniques, the number of features, or the inferential goal (explanation vs prediction) that might contribute to the prediction accuracy of the models. To achieve this, we conduct a meta-analysis of journal articles published in the past 10 years. Based on existing literature, we hypothesize that arousal will be predicted with higher accuracy than valence, as valence tends to be more context-dependent and challenging to model.

<!-- incorporate splits into above? -->

@@

Figure 4 shows forest plot visualization from the random-effects model of the best-performing models in classification studies. $MCC$s vary across a wide range, reporting $MCC$s between 0.55 to 0.98. Table 4 indicates that using the best model from each study increases performance relative to all models ($MCC$ = 0.8684 95%CI: 0.7475-0.9336), yet slightly increases heterogeneity ($\tau^2$ = 0.3180, $I^2$ = 99.8%).

@@

<!-- DISCUSS: 4 digits is too much, let's make this consistent, 2 in all places?, TE 19/3/2025 -->
